---
title: "zAMPExplorer Overview"
author: "Sedreh Nassirnia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zAMPExplorer Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`zAMPExplorer` is an interactive Shiny application for downstream exploration and statistical analysis of 16S rRNA amplicon microbiome data stored in a `phyloseq` object.

Version 0.2.0 uses a modular application architecture. Each analysis tab creates a single analysis/plot object that is reused for the interactive interface, downloads and optional automatic saving. This keeps publication exports consistent with what was inspected in the application.

# Workflow

1. Upload a `.rds` file containing a `phyloseq` object.
2. Inspect sequencing depth and optionally remove low-depth samples in **Read QC**.
3. Explore taxa prevalence, community composition and heatmaps.
4. Compare alpha and beta diversity.
5. Run MaAsLin2, DMM or RDA/dbRDA when these analyses match the study design.
6. Export figures as PDF, SVG or 600-dpi PNG and download result tables.

# Modules

- **Dataset overview**: dataset size, metadata and abundance/taxonomy tables.
- **Read QC**: sequencing-depth plots, filtering and rarefaction.
- **Taxa overview**: prevalence/abundance and shared core taxa.
- **Composition**: relative-abundance stacked barplots.
- **Heatmap**: `ComplexHeatmap` visualization with sample annotations.
- **Alpha diversity**: standard richness/diversity indices and pairwise Wilcoxon tests with BH correction.
- **Beta diversity**: PCA, PCoA and NMDS, plus PERMANOVA and dispersion testing.
- **Differential abundance**: multivariable association testing with MaAsLin2.
- **Community typing**: Dirichlet multinomial mixture models.
- **RDA / dbRDA**: constrained ordination and permutation tests.

# Running the app

```{r, eval=FALSE}
library(zAMPExplorer)
zAMPExplorer_app()
```

During development from a cloned repository:

```{r, eval=FALSE}
devtools::load_all()
zAMPExplorer_app()
```

# Publication outputs

The standard plot controls use dimensions in inches. PDF is recommended for manuscript figures, SVG for editable vector output and PNG for 600-dpi raster output. If automatic saving is enabled, publication PDFs and associated tables are written to the results directory selected on the upload tab.

Statistical settings should always be chosen according to the experimental design. zAMPExplorer provides an interface to the methods; it does not make study-design decisions automatically.
